---
title: "Data manipulation with `dplyr`"
author: "Jingyi Yao"
date: "`r Sys.Date()`"
output: github_document
---
## Preparation Work
### `options()` the global printing settings
#### `options(tibble.print_min = number)` controls how many rows it will show
```{r}
library(tidyverse)
options(tibble.print_min = 3) # it will show at least 3 rows of data

```


### load the litters and puppy data set
#### `clean_names()` delete the blanks and substitute all symbols with `_`
#### `read_csv(col_types="cdi")`controls the variable types
  * `c` for character
  * `d` for doubles
  * `i` for integer
```{r}
litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")      # control the types of variables
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data) # clean names with _ and eliminate other symbols

```


###################################################################################################


## I. `select` columns
#### only need a subset of the columns in a data table

### 1. `select()` fucntion
#### specify the columns you want to keep by **naming** all of them:
#### you need to type the column names
#### the first arugument is the **df name**
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)

```


### 2. specify a *range of columns*
#### do not need to type all the names
#### `group:gd_of_birth` means from `group` to `gd_of_birth`
```{r}
select(litters_data, group:gd_of_birth) 

```


### 3. *remove* columns:
#### put a minus sign `-` before the column name
```{r}
select(litters_data, -pups_survive)

```

### 4. `rename(new_col_name = old_col_name)` the variable *while selecting* variables
#### format: `new_name = old name`
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)

```


### 5. `remame()` function
#### rename but will return all the variables *without selecting*
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)  # a function in tidyverse

```


### 6. `starts_with()`, `ends_with()`, `contains()` in `select()` 
#### using `?select_helpers`
#### when there variables are named with *suffixes* or other *standard patterns*:
```{r}
select(litters_data, starts_with("gd"))    # variable starts with gd
select(litters_data, ends_with("weight"))  # variable ends with weight
select(litters_data, contains("gd"))       # variable contains gd

?select_helpers
```


### 7. `everything()` in `select()`
#### put the selected columns before `everything()`
#### the selected columns will be put at the front
#### the columns not selected will be placed after the selected ones 
```{r}
select(litters_data, litter_number, pups_survive, everything())

```


### 8. `relocate()` *NOT* in `select()` 
#### similar to everything()
#### but `relocate()` is a function outside of `select`
#### the first argument is the df name
```{r}
relocate(litters_data, litter_number, pups_survive)

```




###################################################################################################



## II. `filter()` -- filter out *rows* 
### comparision: `select()` operates on **columns** and `filter()` operates on **rows**

### 1. important operators:
*  comparison operators `>, >=, <, <=, ==, and !=`.
*  `%in%` to detect if values appear in a set
*  `is.na()` to find missing values.
*  logical operators `&` and `|`or negated using `!`


### 2. exmamples:
*  `gd_of_birth == 20`
*  `pups_born_alive >= 2`
*  `pups_survive != 4` not equal to
*  `!(pups_survive == 4)` exclude
*  `group %in% c("Con7", "Con8")` column name in a vector
*  `group == "Con7" & gd_of_birth == 20`


### 3. omit missing observations. 
#### recommend: using `drop_na` from the `tidyr` package:
#### format :  `drop_na(df,colname)`means: drop the rows in df where the column has missing val

*  `drop_na(litters_data)` will remove **any row** with a missing value
*  `drop_na(litters_data, wt_increase)` will remove rows for which `wt_increase` is missing.


### 4. basic examples
```{r}
filter(pups_data, sex == 1)
filter(pups_data, sex == 2, pd_walk < 11)
```




###################################################################################################



## III. `mutate()` *create new variables* to the tibble

### 1. `mutate(new_col = function(old_columns))` add new columns using the old columns

*  Your new variables can be functions of old variables
*  New variables appear **at the end** of the df in the order that they are created
*  You can overwrite old variables
*  You can create a new variable and immediately refer to (or change) it

#### `str_to_lower(col)` change the character all to lower case
```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)          # change capital to lower 
)

```



###################################################################################################



## IV. `arrange()` sort the data
#### format: `arrange(df, col_1,col_2,...)`
#### we may arrange the df by many columns but they have an internla order
#### the default order is ascending, otherwise we need to add `desc(col)` to change it
```{r}
head(arrange(litters_data, group, pups_born_alive), 10) # default : ascending

head(arrange(litters_data, desc(group), pups_born_alive), 10) # change : desc(col_name)
```



###################################################################################################



## V. pipes: `%>%` for *multi-step* data manipulation
### Cmd + Shift + M (Mac) to generate ` %>% `
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```

#### except select,mutate,arrange,filter, we can use `%>%` for other functions 
#### in `lm()`,the argument `data=.`  cannot be left out, we use `.` to represent the data
```{r}
litters_data %>%
  lm(wt_gain ~ pups_born_alive, data = .) %>%    # data = . must be written
  broom::tidy()
```



###################################################################################################



## VI. Learning Check:
#### Learning Assessment: Write a chain of commands that:

*  loads the pups data
*  cleans the variable names
*  filters the data to include only pups with sex 1
*  removes the PD ears variable
*  creates a variable that indicates whether PD pivot is 7 or more days

```{r}
df <- pups_data %>% janitor::clean_names() %>% filter(sex==1) %>% select(- pd_ears)%>% mutate(PD_status = pd_pivot>7) 
df
```

