---
title: "Data manipulation with `dplyr`"
author: "Jingyi Yao"
date: "`r Sys.Date()`"
output: github_document
---

## **`options()`**? the global printing settings?
```{r}
library(tidyverse)

options(tibble.print_min = 3)  # how the data is shown when there are too much data\

```

```{r}
litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")      # control the types of variables
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data) # clean names with _ and eliminate other symbols
```


## **`select`** columns
#### only need a subset of the columns in a data table
#### extracting only what you need can helpfully de-clutter, when you have large datasets. 

### **`select()`** fucntion
#### specify the columns you want to keep by naming all of them:
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)

```

### specify a **range of columns**
#### do not need to type all the names
#### the range has to be continuous? 
```{r}
select(litters_data, group:gd_of_birth) 
```

### **remove** columns:
#### put a minus sign `-` before the column name
```{r}
select(litters_data, -pups_survive)
```

### **rename** the variable while selecting variables
#### format: `new_name = old name`
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

### **`remame()`** function
#### rename but will return all the variables without selecting
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)  # a fucntion in tidyverse

```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```

```{r}


```
