---
title: "Data Import"
output: github_document
---


#### `readr` is in the tidyverse package
```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse) # contains readr

```


## I. Data Import : CSVs

### 1. import data with `readr` package

####  use a *relative* path
####  `folder name/file name.format`  under the r project
####  do not miss the file format like the `.csv`
####  `read_csv()` will guess the type of the columns
####  `read_csv(col_types="ccddii")` change the type when it is automatically wrong
```{r}
litters_df <- read_csv("data/FAS_litters.csv")

``` 



### 2. clean column names with `janitor::clean_names()`

#### do not need to library(janitor) for loading too many things
#### we only need the clean_names funtion in janitor
```{r}
litters_df <- janitor::clean_names(litters_df)  
litters_df

```


### 3. look at the data with `head()` and `tail()`
```{r}
litters_df

head(litters_df)  # first 6 lines
tail(litters_df)
```
#### `head()` shows the first 6 rows
#### `tail()` shows the last 6 rows


### 4. view the whole data frame with `View()`

#### view the whole data set in a new window
#### the result will not be shown in the knitted file


### 5. `skim()` function in the `skimr` package

#### first part of the result is the data summary table
#### then it shows the statistics for each variable(column)
#### statistics include : 
     *  number of missing values
     *  complete rate
     *  mean
     *  sd
     *  quartiles
     *  histogram
```{r}
skimr::skim(litters_df)    

```


### 6. `read_csv` argument options : the parameters in the function

#### for example:
    *  skip 2 rows
    *  change na
    *  change col_types
    
#### `skip= number` how many row to skip
#### `na=c("","NA",".")` means: `""`,`"."` and `"NA"` will be treated as na in R
#### tips: `NA` need the quotation marks `"NA"` to be in a vector
    
```{r,eval=FALSE}
litters_df <- read_csv("data/FAS_litters.csv",na=c("","NA"),skip=2)

```

    


#################################################################################################



##II. Import: *EXCEL* spreadsheet

### 1. ``read_excel` from `readxl` package
```{r}
library(readxl)
mlb_df <- read_excel("data/mlb11.xlsx")

```
#### also using the relative route


### 2. `View()` also applied to excel sheets
```{r,eval=FALSE}
View(mlb_df)
```


### 3. read *PART OF* the table in EXCEL by claiming the *RANGE*
```{r}
Lotr_words_df <- read_excel("data/LotR_Words.xlsx",range="B3:D6")
Lotr_words_df

```


#################################################################################################



## III. Import: *SAS* dataset

### 1. `read_sas` in `haven` package
```{r}
library(haven)
pulse_df <- read_sas("data/public_pulse_data.sas7bdat")
pulse_df

```



#################################################################################################


## IV: Data *Export*

### 1. save the *csv* file in the relative path
#### use `write_csv(name of the file,path)` to export the csv file
```{r}
write_csv(Lotr_words_df,file = "data/lotr_words_df.csv")
```


## why not use the baseR
### we do not use the `read.csv()`
```{r}
dont_do_this_df <- read.csv("data/FAS_litters.csv")
dont_do_this_df
```



